2.4 PRNT50

### 2024-05-28 2_scripts/2.4 PRNT50.qmd
Vax% estimation for PRNT50 (unfinished)

## 1. Clean the APL data
```{r}
APLs<-merge(APLs, select(RFD4682_e, order_ID, 
                   'RBD-IgGII_RESULT', `RBD-IgGII_INTERP`,`RBD-IgGII_RESULT_DATE`
                   ), by="order_ID",all=T)
colnames(APLs)<-c(names(APLs)[1:15], "abbott_s", "pos_abbott_s", "month_abbott_s")
APLs$abb_s_index<-grepl("index", APLs$abbott_s, ignore.case = T)
APLs$abb_s_AU<-grepl("ml", APLs$abbott_s, ignore.case = T)
table(APLs$abb_s_index, APLs$abb_s_AU)
# 36 samples missing abbott_s info
APLs$abb_s_index[APLs$abb_s_index==0 & APLs$abb_s_AU==0]<-NA
APLs$abb_s_AU[is.na(APLs$abb_s_index)]<-NA

@@

# Group 4: no available anti-S data, only Abbott-N
fivenum(APLs[!is.na(APLs$abbott_n) & is.na(APLs$abb_s_AU),]$samplemonth, na.rm = T)
```
**Notes**: So no matter what it really is it can't be Abbott-S-RBD and shall not be included in anti-S analysis.

@@

p09e
```

>>>>

## 1. Clean the APL data

#### 1.1 Use the APLs created
```{r}
APLs<-merge(APLs, select(RFD4682_e, order_ID, 
                   'RBD-IgGII_RESULT', `RBD-IgGII_INTERP`,`RBD-IgGII_RESULT_DATE`
                   ), by="order_ID",all=T)
colnames(APLs)<-c(names(APLs)[1:15], "abbott_s", "pos_abbott_s", "month_abbott_s")
APLs$abb_s_index<-grepl("index", APLs$abbott_s, ignore.case = T)
APLs$abb_s_AU<-grepl("ml", APLs$abbott_s, ignore.case = T)
table(APLs$abb_s_index, APLs$abb_s_AU)
# 36 samples missing abbott_s info
APLs$abb_s_index[APLs$abb_s_index==0 & APLs$abb_s_AU==0]<-NA
APLs$abb_s_AU[is.na(APLs$abb_s_index)]<-NA

@@

# Group 4: no available anti-S data, only Abbott-N
fivenum(APLs[!is.na(APLs$abbott_n) & is.na(APLs$abb_s_AU),]$samplemonth, na.rm = T)
```
**Notes**: So no matter what it really is it can't be Abbott-S-RBD and shall not be included in anti-S analysis. Updates on May 28th: Carmen replied and confirmed they are misplaced anti-N assay results.

#### 1.2 Move the misplaced anti-N back to where they are
```{r}
APLs$abbott_n[APLs$abb_s_index==1 &
              !is.na(APLs$abbott_s)]<-APLs$abbott_s[APLs$abb_s_index==1 & 
                                                   !is.na(APLs$abbott_s)]
APLs$abbott_s[APLs$abb_s_index==1 & !is.na(APLs$abbott_s)]<-NA

# add anti-N interpretation
APLs$pos_abbott[APLs$abbott_n<1.4 & !is.na(APLs$abbott_n)]<-0
APLs$pos_abbott[APLs$abbott_n>=1.4 & !is.na(APLs$abbott_n)]<-1

# redo the anti-S interpretation
APLs$pos_abbott_s<-rep(NA,214780)
APLs$pos_abbott_s[APLs$abbott_s<50 & !is.na(APLs$abbott_s)]<-0
APLs$pos_abbott_s[APLs$abbott_s>=50 & !is.na(APLs$abbott_s)]<-1
table(APLs$pos_abbott_s, useNA = 'ifany')
```

@@

p09e
```
## 4 Misc Analysis

#### 4.1 Sensitivity Analysis
```{r}
# Add vax variable
APLs<-merge(APLs, select(RFD4682_e, order_ID, VACCINATION_STATUS),
      by="order_ID", all=T)

APLs$vax<-APLs$VACCINATION_STATUS
APLs<-select(APLs, -c("VACCINATION_STATUS"))
table(APLs$vax, useNA = 'ifany')

# Add days since imm (vax)
APLs<-merge(APLs, select(RFD4682_e, order_ID, DAYS_SINCE_IMM),
      by="order_ID", all=T)
```
**Notes:** In this part we also do the analysis for those got vaccinated >=28 days, however it does not change the shape of the lines so much. So we just group by doses (0/>=1).

```{r}
# re-do the melt data: no vax
tb_abbsv0<-APLs %>%
  .[.$abb_s_AU==1 & (.$vax %in% c("Not Immunized", "Pre-Immunization")),] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
            )
tb_abbsv0$prop.abbs1<-tb_abbsv0$pos_abbott_s/tb_abbsv0$n
tb_abbsv0$prop.abbs2<-tb_abbsv0$pos_abbott_s2/tb_abbsv0$n

# clean NAs
tb_abbsv0<-tb_abbsv0[!is.na(tb_abbsv0$samplemonth),]

tb_abbsv0_melt<-reshape2::melt(select(tb_abbsv0, -c(2:4)), 
                        id.vars = c("samplemonth"),
                        variable.name = "COI",
                        value.name = "prop")

# re-do the melt data: at least 1 dose
tb_abbsv1<-APLs %>%
  .[.$abb_s_AU==1 & .$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
                             "4 Doses") ,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
            )
tb_abbsv1$prop.abbs1<-tb_abbsv1$pos_abbott_s/tb_abbsv1$n
tb_abbsv1$prop.abbs2<-tb_abbsv1$pos_abbott_s2/tb_abbsv1$n

# clean NAs
tb_abbsv1<-tb_abbsv1[!is.na(tb_abbsv1$samplemonth),]

tb_abbsv1_melt<-reshape2::melt(select(tb_abbsv1, -c(2:4)), 
                        id.vars = c("samplemonth"),
                        variable.name = "COI",
                        value.name = "prop")
```

```{r}
# re-do the plot
# APL: not vaccinated
p09v0<-tb_abbsv0_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
  geom_point(colour="red") +
  geom_line(aes(linetype=COI), colour="red")  +
  theme_bw() + 
  ylim(c(0,1)) +
  xlab("Time of Sample Collection") + 
  ylab("Proportion")+ 
  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
                                                      "APL-Abbott (COI=402)"))+
  theme(legend.position='top') + 
  ggtitle("Stratum: not vaccinated")
p09v0
```
```{r}
# Plot: at least 1 dose
p09v1<-tb_abbsv1_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
  geom_point(colour="red") +
  geom_line(aes(linetype=COI), colour="red")  +
  theme_bw() + 
  ylim(c(0,1)) +
  xlab("Time of Sample Collection") + 
  ylab("Proportion")+ 
  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
                                                      "APL-Abbott (COI=402)"))+
  theme(legend.position='top')+
  ggtitle("Stratum: at least 1 dose")
p09v1
```
#### Figure panel
```{r}
# use cowplot package
p09all.upper<-cowplot::plot_grid(p09c + ylab("Proportion") +
                           theme(legend.position='top') +
                           guides(color = guide_legend(nrow = 2)), 
                           p09v0 + xlab ("") + ylab(""), nrow=1,
                           labels = c("A", "B"),  align = "h")
p09all.upper

p09all.lower<-cowplot::plot_grid(
                           NULL,  p09v1 + theme(legend.position="none"), nrow=1,
                           labels = c("", "C"))
p09all.lower

png(file="../4_output/figs/Fig.4 Immunity.png", 
    width=2400, height=2000, res=300)
cowplot::plot_grid(p09all.upper, p09all.lower, nrow=2, rel_heights = c(1, 0.8))
dev.off()
```

## 4.2 The vaccination rate among APL
The APL data Carmen shared with us has vaccination info, the 2024 Jan ver of it even has the vax brand. While the CBS data we have doesnot include vax info. However, a paper written by Sheila has the vax% of those in CBS data, which we can use for comparison on vax%.

#### 4.2.1 Vax rate by APL vax data
```{r}
# create vax% cat var
APLs$vax_cat<-rep(NA, 214780)
APLs$vax_cat[APLs$vax %in% c("1 Dose","1-2 Doses","2 Doses","3 Doses",
                             "4 Doses")]<-1
APLs$vax_cat[APLs$vax %in% c("Not Immunized", "Pre-Immunization")]<-0
table(APLs$vax, APLs$vax_cat, useNA = 'ifany')
```
```{r}
# create merged data
tb_vax<-APLs %>%
  .[.$abb_s_AU==1,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_vax= sum(vax_cat, na.rm = T)
            )
tb_vax$prop.vax<-tb_vax$n_vax/tb_vax$n

# clean NAs
tb_vax<-tb_vax[!is.na(tb_vax$samplemonth),]

# make the plot
p10<-tb_vax %>%
  ggplot(aes(x=samplemonth, y=prop.vax)) +
  geom_point() +
  geom_line()  +
  geom_smooth(colour="red") +
  theme_bw() + 
  ylim(c(0,1)) +
  ylab("Vaccinated%")
p10
```
**Alton's comments:** "It might be worth comparing to reported vax rate for Alberta similar to what Sheila and Steve did in that figure." As they used (S+ N-)% to estimate the vax%, it will inevitably results in underestimation, as those who are vaccinated and infected was not counted.

#### 4.2.2 Vax by anti-S and anti-N
```{r}
# create an addition indicator for S+ and N-
table(APLs$pos_abbott, APLs$pos_abbott_s, useNA = 'ifany')

APLs$vax_est<-rep(0, 214780)
APLs$vax_est[APLs$pos_abbott==0 & APLs$pos_abbott_s==1]<-1
APLs$vax_est[is.na(APLs$pos_abbott) | is.na(APLs$pos_abbott_s)]<-NA
table(APLs$vax_est)
```
```{r}
tb_vax_est<-APLs %>%
#  .[!is.na(.$pos_abbott_s) & !is.na(.$pos_abbott),] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            n_vax= sum(vax_est, na.rm = T)
            )
tb_vax_est$prop.vax<-tb_vax_est$n_vax/tb_vax_est$n

# clean NAs
tb_vax_est<-tb_vax_est[!is.na(tb_vax_est$samplemonth),]

# make the plot
p11<-tb_vax_est %>%
  ggplot(aes(x=samplemonth, y=prop.vax)) +
  geom_point() +
  geom_line()  +
  geom_smooth(colour="red") +
  theme_bw() + 
  ylim(c(0,1)) +
  ylab("Vaccinated% Est by anti-S&N")
p11
```
**Notes:** Estimating the vax% using the anti-S (+) and anti-N (-) assay results. For our analysis, this way of estimation might be problematic. We don't have pre-Omicron wave anti-N data in APL  and the prevalence of ever infected (S+ & N+) raised sharply in 2022. This will result underestimated vax% in 2022, as presented in the fig above.

#### 4.3.3 The Vax rate in AB general population
```{r}
# load the data from Canada Statistics
vax_AB<-read.csv("https://health-infobase.canada.ca/src/data/covidLive/vaccination-coverage-map.csv")

# create AB subset and clean the data
vax_AB<-vax_AB[vax_AB$prename=="Alberta",]
vax_ABs<-vax_AB %>%
  select(week_end, numtotal_atleast1dose, proptotal_atleast1dose)
colnames(vax_ABs)<-c("weekend", "num.1dose", "prop.1dose")
vax_ABs$weekend<-as.yearmon(vax_ABs$weekend)

# calculate monthly mean prop for vax rate
tb_vax_AB<-vax_ABs %>%
  group_by(weekend) %>%
  summarise(ave.1dose = mean(prop.1dose))

tb_vax_AB<-tb_vax_AB[!is.na(tb_vax_AB$ave.1dose),]
tb_vax_AB$data<-rep("AB_general", length(tb_vax_AB$weekend))

d01<-tb_vax_AB
colnames(d01)<-c("samplemonth", "prop.vax", "data")
```

##### creating a merged data
```{r}
# Estimated vaccination rate: by anti-S + and anti-N neg
View(tb_vax_est)
d02<-select(tb_vax_est, c(1,4))
d02$data<-rep("S_pos_N_Neg", length(d02$samplemonth))

# Estimated vaccination rate: by anti-S only
View(tb_abbs)
d03<-select(tb_abbs, samplemonth, prop.abbs1, study)
colnames(d03)<-c("samplemonth", "prop.vax", "data")
d03$data<-rep("S_pos", length(d03$samplemonth))

# Reported vax rate in APL
View(tb_vax)
d04<-select(tb_vax, c(1,4))
d04$data<-rep("APL_vax", length(d04$samplemonth))

# merge all
dta.vax<-rbind.data.frame(d01, d02, d03, d04)
rm(d01, d02, d03, d04)

dta.vax$prop.vax[dta.vax$data=="AB_general"]<-dta.vax$prop.vax[dta.vax$data=="AB_general"]/100
```

##### Making the plot
```{r}
p12<-dta.vax %>%
  ggplot(aes(x=samplemonth, y=prop.vax, group=data, colour=data)) +
  geom_point(aes(colour=data)) +
  geom_line(aes(colour=data))  +
  theme_bw() + 
  ylim(c(0,1)) +
  ylab("Vaccinated%")
p12
```
